---
title: "Running the preterm-birth prediction pipeline"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Running the preterm-birth prediction pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  # Every chunk in this vignette is shown but NOT executed. A real run of the
  # pipeline takes ~12 minutes and needs Bioconductor packages plus data on disk;
  # building the vignette must stay fast and dependency-light. Console output
  # below is quoted verbatim from an actual run on data/example/.
  eval     = FALSE
)
```

This vignette walks through one complete run of the `ptbpredict` pipeline on the synthetic example
dataset shipped with the package, then shows how to point it at your own data.

**The code chunks are not evaluated when this vignette is built** — a full run takes about twelve
minutes. Everything shown is copy-pasteable, and console output is quoted from a real run.

> **If you arrived from the published article**, read
> [`docs/UPDATE_SINCE_PUBLICATION.md`](https://github.com/martinruhle/Mexican-PretermBirth-analysis/blob/master/docs/UPDATE_SINCE_PUBLICATION.md)
> first. Several methodological updates were made after publication and some reported numbers
> changed.

## What the pipeline does

For each of 12 combinations — 2 models × 3 clinical variable sets × 2 microbiome inputs — it runs a
subject-level nested cross-validation and reports out-of-fold performance:

```
for each of 5 outer folds (subjects, stratified by outcome, seed 123):
    hold out the OUTER-TEST subjects — untouched until the last line

    on the OUTER-TRAINING subjects:
        select taxa            (ANCOM-BC2 on counts, adjusting for age + BMI)
        select clinical vars   (Approach 3 only: univariate screening)
        split them 70/30       -> inner-train / inner-validation

    on INNER-TRAIN:
        fit the recipe         (impute, standardise, dummy-code, filter)
        fit the model          (Random Forest or elastic net)

    on INNER-VALIDATION:
        choose the classification threshold (Youden)

    on OUTER-TEST:
        evaluate — AUROC, PRAUC, sensitivity, specificity, balanced accuracy
```

Note where each step draws its data: **feature selection sees the whole outer-training set, while
the recipe and the model see only inner-train.** Nothing derived from outer-test enters any
selection or fitting decision — that is the property the test suite exists to protect.

## Setup

```{r install}
install.packages("renv")
renv::restore()          # pinned R 4.4.2 library, incl. Bioconductor
```

For interactive work, load the engine from the source tree:

```{r load}
devtools::load_all()
```

The pipeline calls several analysis packages unqualified, inherited from its R Markdown origins, so
attach them before calling engine functions directly (the analysis `.Rmd` and the test setup do
this for you):

```{r attach}
library(phyloseq); library(ANCOMBC); library(vegan); library(zCompositions)
library(pROC); library(PRROC)
library(tidymodels)     # last, so dplyr's verbs win the usual conflicts
```

## The short version

One command runs everything on the synthetic data:

```{r run-all, engine = "bash"}
PTB_PROFILE=example Rscript scripts/run_baseline.R
```

That renders the analysis document and writes three files to `analysis/_output/`:

| File | Content |
|---|---|
| `baseline_postfix.html` | Full report: cohort figures, PCA, feature-selection tables, the 12-combination performance table, ROC/PR curves, feature importance, session info |
| `baseline_postfix_metrics.csv` | The 12 combinations as numbers (mean and SD per metric) |
| `baseline_postfix_cv_summary.rds` | The same summaries as an R object |

`PTB_RUN_TAG=myrun` changes the prefix. The rest of this vignette unpacks what happens inside.

## Step 1 — configuration

Nothing is hardcoded: paths, column names, variable-set definitions, preprocessing parameters, CV
settings and model hyperparameters all come from `config/config.yml`, selected by profile.

```{r config}
cfg <- read_config("example")   # or read_config() -> $PTB_PROFILE, default "default"

cfg$data$matrix_path
#> [1] "data/example/example_genus_rel.csv"
cfg$cv$outer_folds
#> [1] 5
cfg$cv$inner_train_prop
#> [1] 0.7
names(cfg$models)
#> [1] "rf_base"     "glmnet_base"
```

The `example:` profile redefines only the `data:` block and inherits everything else from
`default:`, so the synthetic run uses exactly the same methodology as the real one.

## Step 2 — which columns are microbiome?

By an explicit list in `config/data_dictionary.csv`, never by column position and never by a name
pattern:

```{r dictionary}
taxa <- microbiome_columns(cfg)
length(taxa)
#> [1] 97
head(taxa, 3)
#> [1] "Lactobacillus" "Gardnerella"   "Hydrotalea"
```

Names are read with `check.names = FALSE`, so hyphens and taxonomic prefixes survive intact —
`Escherichia-Shigella` was previously sanitised to `Escherichia.Shigella` on read, which kept it out
of differential-abundance analysis.

## Step 3 — load and validate

```{r load-data}
d <- load_dataset(cfg)          # wide matrix + longitudinal metadata
dim(d$matrix)
#> [1] 113 167

validate_input_data(d$matrix, d$metadata, cfg)
#> ✓ Contrato de datos validado: 97 taxa, 113 muestras, 42 sujetos
#>   (sample_id/subject_id casan con la metadata).
```

`validate_input_data()` is the gate to run before anything else. It checks that required columns
exist, the outcome is binary, every dictionary taxon is present in the matrix, taxa lie in
`[0, 1]`, and the sample and subject keys agree between matrix and metadata. It fails with a
specific message rather than letting a malformed dataset produce plausible-looking numbers.

Split the matrix into its two domains — again by name:

```{r split}
dom <- split_domains(d$matrix, cfg)
names(dom)
#> [1] "microbiome" "clinical"
```

## Step 4 — filtering and the compositional transform

The analysis document applies two filters to the microbiome, in this order:

1. **Prevalence** — keep genera present in ≥5% of samples (`preprocessing$prevalence_filter`).
2. **Contaminants** — drop the genera listed in `preprocessing$contaminant_genera`, common reagent
   contaminants in low-biomass samples.

On the real cohort that is 97 → 59 → 49 genera. The surviving list is what the ANCOM-BC2 branch
searches and what "full microbiome" means.

Relative abundances are compositional and must be transformed before entering a linear model. This
happens in two steps, deliberately separated:

```{r clr}
# 1. Learn per-taxon zero-replacement levels ONCE, over the full microbiome.
#    Outcome-blind by construction: the labels are not an argument.
zero_levels <- fit_clr_zerorepl(micro_genus_full, taxa_cols)

# 2. Apply zero replacement + per-sample CLR wherever it is needed. The levels are
#    passed in, never read from a global.
inner_train <- apply_clr_transform(inner_train, taxa_cols, zero_levels = zero_levels)
inner_val   <- apply_clr_transform(inner_val,   taxa_cols, zero_levels = zero_levels)
outer_test  <- apply_clr_transform(outer_test,  taxa_cols, zero_levels = zero_levels)
```

The CLR is **per sample**: each row is centred by the geometric mean of its own taxa, so every
transformed row sums to zero. Being row-local, it cannot move information between training and test
data — which is why applying it to all three splits is safe.

This is the step updated after publication: a pseudocount of 0.65, appropriate for raw counts, was
added to relative abundances, and after the downstream standardisation step that reduced to a
per-feature rescaling, so the intended compositional transform was not taking effect. See
`docs/UPDATE_SINCE_PUBLICATION.md` §2.1.

## Step 5 — the nested CV loop

Folds are built over **subjects**, not samples, so a subject's longitudinal samples cannot be split
across a train/test boundary:

```{r folds}
set.seed(cfg$cv$seed)                                  # 123
cv_folds <- vfold_cv(subject_labels, v = cfg$cv$outer_folds, strata = preterm)
```

Model specifications come from the config:

```{r specs}
models_spec <- build_model_specs(cfg)
names(models_spec)
#> [1] "rf_base"     "glmnet_base"
```

`rf_base` is `ranger` with 500 trees, `mtry = 4`, `min_n = 10` and impurity importance;
`glmnet_base` is `glmnet` with `penalty = 0.01` and `mixture = 0.5`. Hyperparameters are fixed, not
tuned — with 43 subjects there is no budget for tuning. `build_model_specs()` reads only
`cfg$models`, so the three specifications parked under `cfg$models_future` (`xgboost`, `rpart`,
RF-on-PCA) are ignored.

Then one call per combination:

```{r nested-cv}
result <- train_with_nested_cv(
  model_name          = "glmnet_base",
  model_spec          = models_spec$glmnet_base,
  clinical_data_all   = approach2_clinical_all,
  microbiome_data_all = micro_genus_full,
  approach_name       = "Approach2_Literature",
  microbiome_option   = "ANCOM_Taxa",
  cv_folds            = cv_folds,
  clr_zero_levels     = zero_levels,
  genera_clean        = genera_clean,          # ANCOM_Taxa branch
  abs_data            = load_abs_matrix(cfg)   # ANCOM_Taxa branch: counts
)

result$summary$AUROC_mean
result$fold_results          # one row per outer fold
result$selected_taxa         # taxa chosen per fold (they legitimately differ)
result$test_predictions      # subject-level out-of-fold predictions
```

> **Two caveats when calling the engine directly.** `train_with_nested_cv()` still reads
> `subject_labels` from the calling environment, and its Approach-3 branch calls six univariate
> screening helpers that currently live in the analysis `.Rmd` rather than in `R/`. Both are known
> and flagged in the source. Until they are extracted, the supported way to run Approach 3 is
> through `analysis/integrated_preterm_prediction_workflow.Rmd` (or
> `scripts/run_baseline.R`, which renders it). Approaches 1 and 2 need only `subject_labels`.

`result$test_predictions` is worth knowing about: it holds the out-of-fold predicted probabilities
at subject level, so alternative metrics can be computed from a finished run without re-training
anything. The permutation test uses exactly this to recompute a direction-fixed AUROC.

## Step 6 — reading the output

The 12-combination table is the pipeline's headline output:

```{r results}
m <- read.csv("analysis/_output/baseline_postfix_metrics.csv")
m[order(-m$AUROC_mean), c("Model", "Approach", "Microbiome", "AUROC_mean", "AUROC_sd")]
```

On the **real cohort** the best combination is elastic net / Approach 3 / ANCOM-BC2 taxa at
AUROC 0.760 ± 0.270; the full table is in the repository README.

On the **synthetic example data the numbers are meaningless** and are not expected to resemble the
real ones — different data. The criterion for the example run is only that all 12 combinations and
every figure complete successfully. Treat it as a smoke test of your installation.

## Step 7 — the permutation test

Does the reported discrimination exceed chance? Outcome labels are permuted at the subject level
and the whole nested CV is re-run, 999 times:

```{r permutation, engine = "bash"}
# observed value + per-fold supplementary table (~40 s)
PTB_PERM_MODE=verify Rscript scripts/permutation_test.R

# null distribution: 999 permutations, ~3.6 h on 8 cores
PTB_PERM_MODE=run PTB_PERM_N=999 PTB_PERM_CORES=8 PTB_PERM_TAG=final999 Rscript scripts/permutation_test.R

# p-value, summary tables and figure from the saved null (seconds)
PTB_PERM_MODE=report PTB_PERM_TAG=final999 Rscript scripts/permutation_test.R
```

Result on the real cohort: observed 0.760 against a null of 0.502 ± 0.121, 18 of 998 valid
permutations reaching the observed value, **p = 0.019**. The artefacts are versioned in
`results/permutation_test/`, so none of this needs re-running to be checked.

Two details make the null trustworthy, and both were introduced by the update:

- **The AUROC orientation is fixed** rather than inferred from the data. With an inferred
  orientation the null distribution centred on 0.68 instead of 0.5 and the test had almost no
  power — the same data gave p = 0.139.
- **All three consumers of the outcome are permuted together**: the model's outcome column, the
  stratification labels, and the metadata ANCOM-BC2 uses to select taxa. Permute fewer, and taxa
  are still selected with the true labels — a contaminated null.

## Using your own data

Two files to edit, no changes to `R/`.

**1. Add a profile** to `config/config.yml`, redefining the whole `data:` block (profile
inheritance replaces `data:` wholesale rather than merging key by key):

```yaml
mylab:

  data:
    matrix_path:      "data/mylab/genus_rel.csv"      # relative abundances, rows sum to ~1
    abs_matrix_path:  "data/mylab/genus_counts.csv"   # integer counts, same taxa columns
    metadata_path:    "data/mylab/metadata_long.csv"
    participant_path: null
    data_dictionary:  "config/data_dictionary.csv"
    microbiome_columns_from: "data_dictionary"
```

**2. Fill in `config/data_dictionary.csv`** — one row per column of your matrix, with `role` set to
`microbiome` for taxa and `clinical` for everything else, plus the `subject_id`, `sample_id`,
`outcome` and `gestational_age` keys. Every `role = microbiome` variable must exist, spelled
identically, in both the relative-abundance and the count matrix.

Then:

```{r own-data, engine = "bash"}
PTB_PROFILE=mylab Rscript scripts/run_baseline.R
```

Validate the contract first — it is much cheaper than diagnosing a run that stops midway:

```{r own-validate}
cfg <- read_config("mylab")
d   <- load_dataset(cfg)
validate_input_data(d$matrix, d$metadata, cfg)
```

Note that the default configuration references two clinical variables **by name**:
`edad_cronologicamujer` (maternal age) and `imc_pregestacional` (pre-pregnancy BMI), used in the
ANCOM-BC2 confounder formula and in the Approach 1 and 2 variable lists. Provide columns with those
names, or override the relevant config keys in your profile.

Two known limits, both in exploratory figure code rather than the modelling engine: subject and
sample ids must be integer-like (the PCA figure coerces its join key with `as.integer()`), and every
taxon needs at least two samples with a non-zero value (the zero-replacement method cannot estimate
its hyper-parameter otherwise). Both are described in the repository README.

## Where to go next

| Question | Where |
|---|---|
| What changed after publication, and by how much? | `docs/UPDATE_SINCE_PUBLICATION.md` |
| Current results, run instructions, reusability limits | `README.md` |
| Invariants, tests, how to propose a change | `CONTRIBUTING.md` |
| Permutation test artefacts and how to regenerate them | `results/permutation_test/` |
| Function-level reference | the roxygen blocks in `R/` (`R/io.R`, `R/nested_cv.R`, …) |
| Getting the restricted cohort data | `docs/DATA_ACCESS.md` |
